Clusterization Properties of D-optimal Designs in Infinite Dimensional Bayesian Linear Inverse Problems

### 2023-04-17 latex/intro.tex
get ready to create different branches for different journals

In the process of proving Theorem \ref{thm:char}, we prove a novel
lemma in linear algebra: We show how to decompose a symmetric positive
definite matrix $M$ as $M = AA^t$, where $A$ has unit norm columns
(Lemma \ref{lemma:lax}). We also generalize two lemmas from linear
algebra to infinite-dimensional settings: In Lemma \ref{lemma:MDL} we
prove a matrix determinant lemma in Hilbert spaces, and in Lemma
\ref{lemma:lax} we generalize a lemma \cite{Lax07} for calculating
$\frac{\der}{\der t} \log \det (I + X(t))$, where $X(t)$ is an
operator valued function.

@@

    \end{axis}
  \end{tikzpicture}
  \caption{Sensor clusterization for inferring the initial condition
    of the 1D heat equation. Posterior pointwise standard deviations
    (lines) are plotted over the domain $[0, \pi]$, for varying
    numbers of sensors. Sensor locations (circles) were chosen to
    optimize the D-optimality criterion. Sensor clusterization can be
    clearly seen for six sensors. Only four observation locations are
    used --- two pairs of sensors are so close together they are
    indistinguishable.}
  \label{fig:clusterization illustration}
\end{figure}

@@

Sensor clusterization is illustrated in Figure \ref{fig:clusterization
  illustration}. The posterior covariance does not depend on data, so
the plot has no reference to actual data observed. The posterior
covariance does depend on location of the observation taken, which
were chosen to optimize the D-optimality criterion. Sensor
clusterization is clearly seen for six sensors; it looks like only
four observations were taken. Two pairs of sensors are clustered so
close together that they are indistinguishable from one another.

>>>>

One byproduct of the proof of Theorem \ref{thm:char} is a novel lemma
in linear algebra: We show how to decompose a symmetric positive
definite matrix $M$ as $M = AA^t$, where $A$ has unit norm columns
(Lemma \ref{lemma:lax}). We also generalize two lemmas from linear
algebra to infinite-dimensional settings: In Lemma \ref{lemma:MDL} we
prove a matrix determinant lemma in Hilbert spaces, and in Lemma
\ref{lemma:lax} we generalize a lemma \cite{Lax07} for calculating
$\frac{\der}{\der t} \log \det (I + X(t))$, where $X(t)$ is an
operator valued function.

@@

    \end{axis}
  \end{tikzpicture}
  \caption{Sensor clusterization for inferring the initial condition
    of the 1D heat equation. Posterior pointwise standard deviations
    (lines) are plotted over the domain $[0, \pi]$, for varying
    numbers of sensors. Sensor locations (circles) were chosen to
    optimize the D-optimality criterion. Sensor clusterization occurs
    for six sensors. Only four observation locations are used --- two
    pairs of sensors are so close together they are
    indistinguishable.}
  \label{fig:clusterization illustration}
\end{figure}

@@

Sensor clusterization is illustrated in Figure \ref{fig:clusterization
  illustration}. The posterior covariance does not depend on data, so
the plot has no reference to actual data observed. The posterior
covariance does depend on location of the observation taken, which
were chosen to optimize the D-optimality criterion. Sensor
clusterization is clearly seen for six sensors; it looks like only
four observations were taken. Two pairs of sensors are clustered so
close together that they are indistinguishable from each other.

### 2023-04-17 latex/paper.tex
get ready to create different branches for different journals

\begin{abstract}
  We investigate the problem of sensor clusterization in D-optimal
  experimental design for infinite-dimensional Bayesian linear inverse
  problems. We suggest an analytically tractable model for D-optimal
  designs. We show that sensor clusterization in the case of iid
  observation noise is generic. We also show that model spatially
  correlated errors mitigate the clusterization phenomenon.

@@

>>>>

\begin{abstract}
  We investigate the problem of sensor clusterization in D-optimal
  experimental design for infinite-dimensional Bayesian linear inverse
  problems. Many authors seem to believe that the fact that spatial
  correlations mitigateWe suggest an analytically tractable model for D-optimal
  designs. We argue that sensor clusterization in the case of iid
  observation noise is generic. We also prove that a model error term
  mitigates the clusterization phenomenon.

@@

%% \section{Open Problems}
%% Some open problems to

%% We two phenomena for a generic inverse problem with unit-norm
%% constraints on obsevations: First, we showed that when model error is
%% present and obsevation noise is small --- sensor clusterization will
%% not occur. In contrast, we saw that when no model error is present ---
%% sensor clusterization is quite generic and poses no obstruction to
%% D-optimality.

### 2023-04-17 latex/prelims.tex
get ready to create different branches for different journals

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectfully). Denote
$\hilo^* \simeq \hilo$ the Hilbert space of all linear functionals on
$\hilo$. Let $\fwd: \hilp \to \hilo$ the \emph{forward operator}. The
forward operator $\fwd$ is assumed linear and strongly smoothing (has
fast decaying modes --- the heat operator from section
\ref{subsec:example} is a prime example). Take a Gaussian prior
$\param \sim \pr = \normal(\prmean ,\prcov)$ with some appropriate
covariance operator $\prcov$ \cite{Stuart10} on $\hilp$. Note that
$\fwd \prcov \fwd^*$ is the prior covariance in $\hilo$
\cite{Stuart10}. As such, it is assumed invertible --- an assumption
which will be used heavily below. Let $m$ the number of observations
taken. Observations are taken via the \emph{observation
operator}\footnote{It is common for the observation and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires $\fwd$ and $\obs$ are explicitly
separated.} $\obs \in ( \hilo^* )^m$. In an analogy with linear
algebra, where row vectors are thought of as linear functionals,
$\obs$ is best thought of as having ``rows'' $\meas_j, j=1,\dots,m$:
\begin{equation}\label{eq:O}
  \obs = (\meas_1,\dots, \meas_m)^t, \meas_j \in \hilo^*, j = 1,\dots,m.
\end{equation}
This way, $\obs u = (\meas_1(u), \dots, \meas_m(u) )^t \in \R^m$ and
$u \in \hilo$.
%% but we may drop the parentheses and write $\meas_j u = \meas_j(u)$.
%% A few observations regarding $\obs$ are in order.
%% First, it is good to keep in mind that $( \hilo^* )^m$ is a Hilbert
%% space with norm $\| \obs \| = \sum_{j=1}^m \|\meas_j\|$.
For $u \in \hilo$ and $v\in \R^m$:
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

%% \begin{observation}
%%   It is best to think of $\meas_j,j=1,\dots,m$ as row vectors and of
%%   $\obs$ as a matrix with rows $\meas_j$. 
%% \end{observation}

Each observation vector is a linear functional, chosen from (some
subset of) $\hilo^{*}$. Data is acquired via noisy observations, with
two types of error terms considered. Spatially correlated model error
is denoted $\eps' \sim \normal(0,\modcov)$, and modeled as a centered
Gaussian measure on $\hilo$ with covariance operator
$\modcov$. Observation error is denoted $\eps \sim \normal(0, \sigma^2
I_m)$, with $I_m \in \mathbb{R}^{m \times m}$ the identity. Both error
terms and the prior are assumed independent of each other. Thus, data
is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}
%% \subsubsection{Error Terms}
It is easy to verify that $\obs \eps' + \eps \in \R^m$ is a centered
Gaussian random vector with covariance matrix

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j \\
    %
    %
    %
    &= \meas_i (\modcov \meas_j) \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
%% and $e_j \in \R^m$ is the $j$th standard basis vector.
%% The explicit dependence on $\obs$ will be mostly dropped for
%% convenience, so $\Sigma(\obs) = \Sigma$.  Thus, for a fixed $\obs$
%% (i.e.\ a fixed set of observations), $\data = \obs \fwd \param +
%% \bar{\eps}$ with $\bar{\eps} \sim \normal(0,\Sigma)$.
Taking $\modcov = 0$ is a common practice
\cite{Tarantola05,Kaipio2006,Vogel02} and then $\Sigma = \sigma^2I$ is
simply a scalar matrix that does not depend on $\obs$.

@@

Finally, it is useful to record a known result regarding the posterior
covariance operator:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

\subsection{D-Optimal Designs in Infinite Dimensions}\label{subsec:D optimal design} 
A D-optimal design maximizes expected KL divergence between posterior
and prior measures, denoted $\post$ and $\pr$, respecively. It is
useful to recall the definition of KL divergence for an arbitrary
prior measure:
$$
KL(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
$$

D-optimal designs in infinite-dimensional Hilbert spaces were
investigated in \cite{AlexanderianGloorGhattas14}. In
infinite-dimensions, a D-optimal design is well-defined as maximizing
the expected KL divergence between posterior and prior. The main
result is summarized, in the notation presented above, in the
following theorem\footnote{In \cite{AlexanderianGloorGhattas14},
results are stated for $\Sigma=I$ (implied by $\modcov =
0,\sigma^2=1$), but the authors also mention these results hold for
more general covariance matrices (p. 681).}:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where rows of $\obs$ are in some allowed
  set of measurement.
\end{definition}

\noindent In order to undestand the intuition behind Definition
\ref{def:d optimality}, note that from \eqref{eq:postcov} and
\eqref{eq:objective}:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \\
    &= \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \det \prcov \det \postcov^{-1}.
  \end{split}
\end{align*}
Since $\prcov$ is constant, a D-optimal design minimizes the posterior
covariance determinant, analogously to the finite-dimensional case.

>>>>

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectfully). Denote
$\hilo^* \simeq \hilo$ the Hilbert space of all linear functionals on
$\hilo$. Let $\fwd: \hilp \to \hilo$ the \emph{forward operator}. The
forward operator $\fwd$ is assumed linear and strongly smoothing (has
fast decaying modes --- the heat operator from section
\ref{subsec:example} is a prime example). Take a Gaussian prior
$\param \sim \pr = \normal(\prmean ,\prcov)$ with some appropriate
covariance operator $\prcov$ \cite{Stuart10} on $\hilp$. Note that
$\fwd \prcov \fwd^*$ is the prior covariance in $\hilo$
\cite{Stuart10}. As such, $\fwd \prcov \fwd^*$ is assumed invertible
--- an assumption which will be used below. Let $m$ the number of
observations taken. Observations are taken via the \emph{observation
operator}\footnote{It is common for the observation and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires $\fwd$ and $\obs$ are explicitly
separated.} $\obs \in ( \hilo^* )^m$, so that:

\begin{equation*}%\label{eq:O}
  \obs u = (\meas_1(u), \dots, \meas_m(u) )^t \in \R^m,\ u \in \hilo,
\end{equation*}
%
%
and we refer to $\meas_j, j=1,\dots,m$ as \emph{observation
vector}s. For $u \in \hilo$ and $v\in \R^m$:
%
%
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error is denoted $\eps' \sim
\normal(0,\modcov)$, and modeled as a centered Gaussian measure on
$\hilo$ with covariance operator $\modcov$. Observation error is
denoted $\eps \sim \normal(0, \sigma^2 I_m)$, with $I_m \in
\mathbb{R}^{m \times m}$ the identity. Both error terms and the prior
are assumed independent of each other. Thus, data is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}
%% \subsubsection{Error Terms}
It is easy to verify that $\obs \eps' + \eps \in \R^m$ is a centered
Gaussian random vector with covariance matrix

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j \\
    %
    %
    %
    &= \meas_i (\modcov \meas_j) \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
%% and $e_j \in \R^m$ is the $j$th standard basis vector.
%% The explicit dependence on $\obs$ will be mostly dropped for
%% convenience, so $\Sigma(\obs) = \Sigma$.  Thus, for a fixed $\obs$
%% (i.e.\ a fixed set of observations), $\data = \obs \fwd \param +
%% \bar{\eps}$ with $\bar{\eps} \sim \normal(0,\Sigma)$.
Taking $\modcov = 0$ is a common practice
\cite{Tarantola05,Kaipio2006,Vogel02} and then $\Sigma = \sigma^2I_m$
is a scalar matrix and does not depend on $\obs$.

@@

The posterior measure $\post$ is Gaussian in this setting and its
covariance operator is:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

\subsection{D-Optimal Designs in Infinite Dimensions}\label{subsec:D optimal design} 
A D-optimal design maximizes expected KL divergence between posterior
and prior measures, $\post$ and $\pr$. It is useful to first recall
the definition of KL divergence:% for an arbitrary prior measure:
$$
D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
$$

D-optimal designs in infinite-dimensional Hilbert spaces were
investigated in \cite{AlexanderianGloorGhattas14}. Their main result
is summarized, in the notation presented above, in the following
theorem\footnote{In \cite{AlexanderianGloorGhattas14}, results are
stated for $\Sigma=I$ (implied by $\modcov = 0,\sigma^2=1$), but the
authors also mention these results hold for more general covariance
matrices (p. 681).}:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$ are constrained to
  some allowed set of measurement vectors in $\hilo^*$.
\end{definition}

\noindent For an intuitive understanding of Definition \ref{def:d
  optimality}, note that from \eqref{eq:postcov} and
\eqref{eq:objective}:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \\
    &= \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \det \prcov \det \postcov^{-1}.
  \end{split}
\end{align*}
Since $\prcov$ is constant, a D-optimal design minimizes the posterior
covariance determinant, analogously to the finite-dimensional case.
